### 2020-04-26 ncov.tex
edits

We present a statistical framework for averaging across estimated basic reproductive number \Ro from multiple studies and apply the method to seven disparate models published online as pre-prints between January 23--26, 2020 that estimated \Ro for the SARS-CoV-2 outbreak \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
Previous studies have directly calculated the average of reported \Ro values but such methods undermine differences in underlying model assumptions and statistical methods \citep{majumder2020early, liu2020reproductive}.
Instead, we decompose estimated \Ro into three key quantities ($r$, $\bar G$, and $\kappa$) and calculate the average (pooled estimates) of these key parameters.
Calculating \Ro based on the pooled estimates allow us to appropriately average across the uncertainties present in modeling approaches and their underlying assumptions.
We use these pooled estimates to illustrate the importance of propagating different sources of error, particularly uncertainty in both the growth rate and the generation interval.
We also use our framework to tease apart which assumptions of these different models led to their different estimates and the associated degrees of uncertainty.
Despite the availability of more recent and/or updated estimates of \Ro, we restrict ourselves to the estimates above in order to focus on the resolution of uncertainty in the earliest stages of an epidemic.

@@

Despite a wide range of models considered across Study 1--7, all of them assume that the epidemic grows exponentially in the beginning.
While the IDEA model (used in Study 7) includes a discount parameter $d$ that allows the model to deviate from the exponential growth when $d \neq 0$ \citep{fisman2013idea}, Study 7 estimates $d=0$ across all parameters they consider.
When the epidemic is growing exponentially, the estimated basic reproductive number entirely depends on the exponential growth rate $r$ and the \emph{intrinsic} generation-interval distribution $g(\tau)$ \citep{wallinga2007generation}:
\begin{linenomath*}
\begin{equation}
\frac{1}{\Ro} = \int \exp(-r\tau) g(\tau) \, d\tau.
\label{eq:euler}
\end{equation}
\end{linenomath*}
In other words, all model assumptions come down to assumptions about the exponential growth rate $r$ and the shape of the generation-interval distribution $g(\tau)$.
For example, if a model relies on strong assumptions about the underlying observation or process model, the estimated confidence intervals associated with the exponential growth rates or parameters of the generation-interval distributions will be necessarily narrow.
Therefore, it is sufficient to consider the estimates and assumptions about the exponential growth rates and the shapes of the generation-interval distributions to understand disparate estimates of the basic reproduction number.

@@

\subsection{Statistical framework}

@@

As most studies do not report their estimates of the exponential growth rate, we first re-estimate the exponential growth rate that correspond to their model assumptions:
Since the estimates of the basic reproductive number \Ro entirely depends on the exponential growth rate and the shape of generation-interval distributions, we can calulate the exponential growth rate from the basic reproductive number \Ro, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$.
First, to account for uncertainties in these parameters, we model reported values of the basic reproductive number \Ro, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$ with appropriate probability distributions;
we used gamma distributions to model values reported with confidence/credible intervals and uniform distributions to model values reported with ranges (Table 2).
For example, Study 3 estimated $\Ro = 2.92$ (95\% CI: 2.28--3.67);
we model this estimate as a gamma distribution with a mean of 2.92 and a shape parameter of 67, which has a 95\% probability of containing a value between 2.28 and 3.67 (see Table 2 for a complete description).

For each study $i$, we construct a family of parameter sets by drawing 100,000 random samples from the probability distributions (Table 2) that represent the estimates of \Rx{0i} and the assumed values of $\bar G_i$ and $\kappa_i$ and calculate the exponential growth rate $r_i$ via the inverse of \eref{gamma}:
\begin{linenomath*}
\begin{equation}
r_i = \frac{\Rx{0i}^{\kappa_i} - 1}{\kappa_i \bar{G}_i}.
\end{equation}
\end{linenomath*}
This allows us to approximate the probability distributions of the estimated exponential growth rates by each study;
uncertainties in the probability distributions that we calculate for the estimated exponential growth rates will reflect the methods and assumptions that the studies rely on.

\newcommand{\gammdist}{\mathrm{Gamma}}
\begin{table}[t]
\begin{center}
\scriptsize
\begin{tabular}{l|p{4.5cm}|p{2.5cm}|p{2.7cm}}
 & Basic reproductive\newline number \Ro & Mean generation\newline interval $\bar G$ (days) & Generation-interval\newline dispersion $\kappa$ \\
\hline
Study 1 & $\mathrm{Uniform}(1.5, 3.5)$ & 10 & 1 \\
\hline
Study 2 & $\gammdist(\mathrm{mean}=2.6, \alpha=28)$ & 8.4 & 0.5 \\
\hline
Study 3 & $\gammdist(\mathrm{mean}=2.92, \alpha=67)$ & 8.4 & 0.2 \\
\hline
Study 4 & $\gammdist(\mathrm{mean}=3.8, \alpha=1400)$ & 7.6 & 0.5 \\
\hline
Study 5 & $\gammdist(\mathrm{mean}=2.2, \alpha=12)$ & $\mathrm{Uniform}(7, 14)$ & 0.5\\
\hline
Study 6 & $\gammdist(\mathrm{mean}=5.47, \alpha=54)$ & $\mathrm{Uniform}(7.6, 8.4)^\ast$ & 0.2\\
\hline
Study 7 & $\exp(r \bar G)^\dagger$ & $\mathrm{Uniform}(6, 10)$ & 0\\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Probability distributions for \Ro, $\bar G$, and $\kappa$.}
We use these probability distributions to obtain a probability distribution for the exponential growth rate $r$.
The gamma distribution is parameterized by its mean and shape $\alpha$.
Constant values are fixed according to Table 1.
$^\ast$We do not account for this uncertainty during our recalculation of the exponential growth rate $r$ because the reported estimate of $\mathcal R_0$ and its uncertainty assumes $\bar G = 8$; the original article reports three $\mathcal R_0$ (and 95\% CIs) estimates using three different values of $\bar G$: 7.6 (MERS-like), 8 (average), and 8.4 (SARS-like).
We still account for this uncertainty in our pooled estimates ($\mu_G$).
$^\dagger$Study 6 uses the IDEA model \citep{fisman2013idea}, through which the authors effectively fit an exponential curve to the cumulative number of confirmed cases without propagating any statistical uncertainty.
Instead of modeling \Ro with a probability distribution and recalculating $r$, we use $r=0.114\,\mathrm{days}^{-1}$, which explains all uncertainty in the reported \Ro, when combined with the considered range of $\bar G$.
}
\end{table}

>>>>

We present a statistical framework for averaging across estimates of basic reproductive number \Ro from multiple studies and apply the method to seven disparate models published online as pre-prints between January 23--26, 2020 that estimated \Ro for the SARS-CoV-2 outbreak \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
Previous studies have directly calculated the average of reported \Ro values but such methods undermine differences in underlying model assumptions and statistical methods \citep{majumder2020early, liu2020reproductive}.
Instead, we decompose estimated \Ro into three key quantities ($r$, $\bar G$, and $\kappa$) and calculate the average (pooled estimates) of these key parameters.
Calculating \Ro based on the pooled estimates allow us to appropriately average across the uncertainties present in modeling approaches and their underlying assumptions.
We use these pooled estimates to illustrate the importance of propagating different sources of error, particularly uncertainty in both the growth rate and the generation interval.
We also use our framework to tease apart which assumptions of these different models led to their different estimates and the associated degrees of uncertainty.
Despite the availability of more recent and/or updated estimates of \Ro, we restrict ourselves to the estimates above in order to focus on the resolution of uncertainty in the earliest stages of an epidemic.

@@

Despite a wide range of models considered across Study 1--7, all of them assume that the epidemic grows exponentially in the beginning.
While the IDEA model (used in Study 7) includes a discount parameter $d$ that allows the model to deviate from the exponential growth when $d \neq 0$ \citep{fisman2013idea}, Study 7 estimates $d=0$ across all parameters they consider.
When the epidemic is growing exponentially, the estimated basic reproductive number entirely depends on the exponential growth rate $r$ and the \emph{intrinsic} generation-interval distribution $g(\tau)$ \citep{wallinga2007generation}:
\begin{linenomath*}
\begin{equation}
\frac{1}{\Ro} = \int \exp(-r\tau) g(\tau) \, d\tau.
\label{eq:euler}
\end{equation}
\end{linenomath*}
In other words, all model assumptions come down to assumptions about the exponential growth rate $r$ and the shape of the generation-interval distribution $g(\tau)$.
For example, if a model relies on strong assumptions about the underlying observation or process model, the estimated confidence intervals associated with the exponential growth rates or parameters of the generation-interval distributions will be necessarily narrow.
Therefore, it is sufficient to consider the estimates and assumptions about the exponential growth rates and the shapes of the generation-interval distributions to understand disparate estimates of the basic reproductive number.

@@

\subsection{Re-estimation of the exponential growth rate}

As most studies do not report their estimates of the exponential growth rate, we first re-estimate the exponential growth rate that correspond to their model assumptions.
Since the estimates of the basic reproductive number \Ro entirely depends on the exponential growth rate and the shape of generation-interval distributions, we can calulate the exponential growth rate from the basic reproductive number \Ro, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$.
First, to account for uncertainties in these parameters, we model reported values of the basic reproductive number \Ro, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$ with appropriate probability distributions.
We used gamma distributions to model values reported with confidence/credible intervals and uniform distributions to model values reported with ranges;
when confidence/credible intervals are reported, we parameterize the gamma distribution such that (i) its mean matches the estimate value and (ii) the probability that a random variable following the specified gamma distribution falls between the lower and upper confidence/credible limits is equal to the reported confidence/credible level. This probability does not need to be equi-tailed.
For example, Study 3 estimated $\Ro = 2.92$ (95\% CI: 2.28--3.67);
we model this estimate as a gamma distribution with a mean of 2.92 and a shape parameter of 67, which has a 95\% probability of containing a value between 2.28 and 3.67 (see Table 2 for a complete description).

For each study $i$, we construct a family of parameter sets by drawing $10^5$ random samples from the probability distributions (Table 2) that represent the estimates of $(\Ro)_{i,m}$ and the assumed values of $\bar G_{i,m}$ and $\kappa_{i,m}$ and calculate the exponential growth rate $r_{i,m}$ via the inverse of \eref{gamma}:
\begin{linenomath*}
\begin{equation}
r_{i,m} = \frac{\left[(\Ro)_{i,m}\right]^{\kappa_{i,m}} - 1}{\kappa_{i,m} \bar{G}_{i,m}},
\end{equation}
\end{linenomath*}
where $m=1,\dots,10^5$.
This allows us to approximate the probability distributions of the estimated exponential growth rates by each study.
Uncertainties in the probability distributions that we calculate for the estimated exponential growth rates will reflect the methods and assumptions that the studies rely on.
Then, the set $(\bar{G}_{i}, \kappa_{i}, r_{i})$ defines an empirical parameter distribution for each study without having to re-fit each model.

For most cases, we model $(\Ro)_i$, $\bar G_i$, and $\kappa_i$ using independent distributions (as specified in Table 2) to calculate $r_i$.
In practice, this assumption does not affect our analysis because we are re-calculating the probability distribution of $r_i$ that is consistent with the reported values of $(\Ro)_i$, $\bar G_i$, and $\kappa_i$;
in other words, we still obtain the same estimates and the associated uncertainty of $(\Ro)_i$ if we calculate it from $r_i$, $\bar G_i$, and $\kappa_i$.
For study 6, we fix $\bar G=8\,\textrm{days}$ and use the gamma distribution (Table 2) that corresponds to $\mathcal R_0 = 5.47$ (95\% CI: 4.16--7.10) during the re-estimation step for $r$ to remain consistent with the original study that assumed $\bar G=8\,\textrm{days}$ to obtain this particular estimate;
the original article reports three $\mathcal R_0$ (and 95\% CIs) estimates using three different values of $\bar G$: 7.6 (MERS-like), 8 (average), and 8.4 (SARS-like) days.
We account for uncertainties in $\bar G$ for Study 6 (Table 1) in all other steps in order to properly incorporate parameter uncertainties in the estimate of \Ro.
Study 7 uses the IDEA model \citep{fisman2013idea}, through which the authors effectively fit an exponential curve to the cumulative number of confirmed cases without propagating any statistical uncertainty.
Instead of modeling \Ro with a probability distribution and recalculating $r$, we use $r=0.114\,\mathrm{days}^{-1}$, which explains all uncertainty in the reported \Ro, when combined with the considered range of $\bar G$ in the original article.

\newcommand{\gammdist}{\mathrm{Gamma}}
\begin{table}[t]
\begin{center}
\scriptsize
\begin{tabular}{l|p{4.5cm}|p{2.5cm}|p{2.7cm}}
 & Basic reproductive\newline number \Ro & Mean generation\newline interval $\bar G$ (days) & Generation-interval\newline dispersion $\kappa$ \\
\hline
Study 1 & $\mathrm{Uniform}(1.5, 3.5)$ & 10 & 1 \\
\hline
Study 2 & $\gammdist(\mathrm{mean}=2.6, \alpha=28)$ & 8.4 & 0.5 \\
\hline
Study 3 & $\gammdist(\mathrm{mean}=2.92, \alpha=67)$ & 8.4 & 0.2 \\
\hline
Study 4 & $\gammdist(\mathrm{mean}=3.8, \alpha=1400)$ & 7.6 & 0.5 \\
\hline
Study 5 & $\gammdist(\mathrm{mean}=2.2, \alpha=12)$ & $\mathrm{Uniform}(7, 14)$ & 0.5\\
\hline
Study 6 & $\gammdist(\mathrm{mean}=5.47, \alpha=54)$ & $\mathrm{Uniform}(7.6, 8.4)^\ast$ & 0.2\\
\hline
Study 7 & $\exp(r \bar G)^\dagger$ & $\mathrm{Uniform}(6, 10)$ & 0\\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Probability distributions for \Ro, $\bar G$, and $\kappa$.}
We use these probability distributions to obtain a probability distribution for the exponential growth rate $r$.
The gamma distribution is parameterized by its mean and shape $\alpha$.
Constant values are fixed according to Table 1.
$^\ast$We do not account for this uncertainty during our recalculation of the exponential growth rate $r$ because the reported estimate of $\mathcal R_0$ and its uncertainty assumes $\bar G = 8$; the original article reports three $\mathcal R_0$ (and 95\% CIs) estimates using three different values of $\bar G$: 7.6 (MERS-like), 8 (average), and 8.4 (SARS-like).
We still account for this uncertainty in our pooled estimates ($\mu_G$).
$^\dagger$Study 7 uses the IDEA model \citep{fisman2013idea}, through which the authors effectively fit an exponential curve to the cumulative number of confirmed cases without propagating any statistical uncertainty.
Instead of modeling \Ro with a probability distribution and recalculating $r$, we use $r=0.114\,\mathrm{days}^{-1}$, which explains all uncertainty in the reported \Ro, when combined with the considered range of $\bar G$.
}
\end{table}

\subsection{Pooled-estimates}

@@
